---
title: "MSstats PSM-Level Proteomics Analysis"
lightbox: true
subtitle: "lionakism-20250630-NPW1 example dataset excluding sample F22"
author:
  - Kevin Cormier
date: "`r Sys.Date()`"
title-block-banner: "#27445C"
format:
  html:
    toc: true
    toc-location: left
    toc-active-border: red
    page-layout: full
    theme:
      light: cosmo
      dark: darkly
    html-math-method: katex
    code-fold: true
    highlight-style: tango
    toc_float: true
    number-sections: false
    embed-resources: true
    smooth-scroll: true
    fig-align: 'center'
    fig-width: 10
    fig-height: 10
css: style.css
editor: source
execute:
  cache: false
editor_options:
  chunk_output_type: console
params:
  request: "lionakism-20250630-NPW1-exclude-F22"
  psm_file: "../data/lionakism-20250630_PSM.xlsx"
  annotation_file: "../data/lionakism-20250630_Annotation.xlsx"
  output_dir: "../results/lionakism-20250630-NPW1-exclude-F22"
  max_psm_rows: null
  excluded_runs:
    - "F22"
  filter_confidence: "High"
  remove_contaminants: true
  remove_shared_peptides: true
  remove_single_peptide_proteins: false
  remove_few_measurements: false
  normalization: "equalizeMedians"
  summary_method: "TMP"
  mb_impute: true
  number_of_cores: 1
  volcano_p_cutoff: 0.05
  volcano_fc_cutoff: 1
  volcano_top_labels: 10
  volcano_split_axis_quantile: 0.95
  volcano_split_axis_min_ratio: 1.8
  volcano_split_axis_scales: 0.35
  qc_max_features: 5000
  qc_top_cv_proteins: 25
  power_alpha: 0.05
  power_target: 0.8
  power_effect_log2fc: 1
  power_top_rows: 200
  gsea_pvalue_cutoff: 0.05
  gsea_min_gene_set_size: 100
  gsea_max_gene_set_size: 500
  gsea_dot_show_categories: 10
  gsea_tree_show_categories: 20
  gsea_tree_label_wrap: 40
---

```{r, include=FALSE}
if (.Platform$OS.type == "windows") {
  options(download.file.method = "wininet")
}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      cache = FALSE, fig.align = 'center',
                      fig.width = 10, fig.height = 10,
                      comment = NA)
```

```{r setup}
required_packages <- c(
  "MSstats",
  "readxl",
  "dplyr",
  "tidyr",
  "stringr",
  "purrr",
  "ggplot2",
  "ggrepel",
  "scales",
  "EnhancedVolcano",
  "ggbreak",
  "readr",
  "DT",
  "knitr",
  "tibble",
  "clusterProfiler",
  "AnnotationDbi",
  "org.Hs.eg.db",
  "enrichplot",
  "htmltools"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop("Install required package(s): ", paste(missing_packages, collapse = ", "), call. = FALSE)
}

suppressPackageStartupMessages({
  library(MSstats)
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(purrr)
  library(ggplot2)
  library(EnhancedVolcano)
  library(ggbreak)
  library(readr)
  library(DT)
  library(knitr)
  library(tibble)
})

current_input <- tryCatch(knitr::current_input(), error = function(e) NULL)
report_dir <- if (!is.null(current_input) && nzchar(current_input)) {
  dirname(normalizePath(current_input, winslash = "/", mustWork = TRUE))
} else {
  normalizePath(getwd(), winslash = "/", mustWork = TRUE)
}

psm_file <- normalizePath(file.path(params$psm_file), winslash = "/", mustWork = TRUE)
annotation_file <- normalizePath(file.path(report_dir, params$annotation_file), winslash = "/", mustWork = TRUE)
output_dir <- file.path(report_dir, params$output_dir)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
read_psm_n_max <- if (is.null(params$max_psm_rows)) Inf else params$max_psm_rows
source(file.path(report_dir, "pathway-analysis-functions.R"), local = knitr::knit_global())
source(file.path(report_dir, "qc-helper-functions.R"), local = knitr::knit_global())

figure_dir <- file.path(report_dir, "..", "figures")
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
knitr::opts_chunk$set(
  fig.path = paste0("../figures/", params$request, "-"),
  dev = c("png", "pdf"),
  dpi = 150
)

set.seed(20250707)
```

## Methods

The Protein Spectral Match (PSM-level) reporter-ion intensities were imported from the Proteome Discoverer
export and linked to experimental condition, biological replicate, and raw-file
metadata. The F22 run and its annotation record were excluded before quality control and statistical modeling. Quantitative
rows with missing or non-positive abundance were removed; when available, only
`r params$filter_confidence`-confidence PSMs were retained andcontaminant-flagged rows were excluded. Peptides shared among proteins were
removed during conversion to MSstats, while proteins represented by a single
peptide were retained.

Proteome Discoverer fields were converted with
`MSstats::PDtoMSstatsFormat()`, resolving duplicate feature measurements by
their maximum intensity. Intensities were log2 transformed, normalized by
`r params$normalization`, summarized from peptide features to proteins with the
`r params$summary_method` method, and missing values were handled using
MSstats model-based imputation. Quality assessment included run-level feature
counts and intensity distributions, missingness, coefficients of variation,
Pearson correlations, and principal component analysis.

Protein-level linear models tested FC versus Healthy, LC versus Healthy, and LC
versus FC. The Infected-versus-Healthy contrast represented the equally
weighted mean of FC and LC minus Healthy. MSstats supplied log2 fold changes,
standard errors, test statistics, degrees of freedom, raw p-values, and
Benjamini-Hochberg adjusted p-values. Differential expression was defined as
an adjusted p-value below `r params$volcano_p_cutoff` and an absolute log2 fold
change of at least `r params$volcano_fc_cutoff`.

For GO Biological Process gene set enrichment analysis (GSEA), UniProt protein
accessions were mapped to Entrez identifiers and ranked by signed statistical
evidence. Gene sets containing `r params$gsea_min_gene_set_size` to `r params$gsea_max_gene_set_size` 
genes were tested with clusterProfiler; results
were interpreted using Benjamini-Hochberg adjusted p-values and summarized in
dot plots, similarity tree plots, and result tables. Power diagnostics used a
two-sided alpha of `r params$power_alpha`, target power of `r params$power_target`, and target absolute effect of `r params$power_effect_log2fc` log2 units.


```{r load-data}
psm_raw <- readxl::read_excel(
  psm_file,
  sheet = "PSMs",
  .name_repair = "minimal",
  n_max = read_psm_n_max
)

annotation_raw <- readxl::read_excel(annotation_file, .name_repair = "minimal")

required_psm_cols <- c(
  "File ID",
  "Annotated Sequence",
  "Charge",
  "Master Protein Accessions",
  "Quan Value"
)
required_annotation_cols <- c("File Name", "File ID", "Condition", "BioReplicate")

missing_psm_cols <- setdiff(required_psm_cols, names(psm_raw))
missing_annotation_cols <- setdiff(required_annotation_cols, names(annotation_raw))
if (length(missing_psm_cols) > 0) {
  stop("Missing required PSM column(s): ", paste(missing_psm_cols, collapse = "; "), call. = FALSE)
}
if (length(missing_annotation_cols) > 0) {
  stop("Missing required annotation column(s): ", paste(missing_annotation_cols, collapse = "; "), call. = FALSE)
}

psm <- psm_raw |>
  rename(SpectrumFile = `File ID`) |>
  mutate(
    SpectrumFile = as.character(SpectrumFile),
    `Quan Value` = suppressWarnings(as.numeric(`Quan Value`))
  )

annotation <- annotation_raw |>
  rename(Run = `File ID`) |>
  mutate(
    Run = as.character(Run),
    Condition = as.character(Condition),
    BioReplicate = as.character(BioReplicate),
    SampleName = tools::file_path_sans_ext(basename(as.character(`File Name`))),
    FigureLabel = .data$SampleName
  )

excluded_runs <- as.character(params$excluded_runs)
excluded_runs <- excluded_runs[!is.na(excluded_runs) & nzchar(excluded_runs)]
if (length(excluded_runs) > 0) {
  missing_excluded_runs <- setdiff(excluded_runs, annotation$Run)
  if (length(missing_excluded_runs) > 0) {
    warning("Excluded run(s) were not found in annotation: ",
            paste(missing_excluded_runs, collapse = ", "))
  }

  psm <- psm |>
    filter(!.data$SpectrumFile %in% excluded_runs)
  annotation <- annotation |>
    filter(!.data$Run %in% excluded_runs)
}

missing_runs <- setdiff(unique(psm$SpectrumFile), annotation$Run)
if (length(missing_runs) > 0) {
  stop("Missing annotation for run(s): ", paste(missing_runs, collapse = ", "), call. = FALSE)
}
```

```{r qc-helper-functions}
as_plain_tibble <- function(x) {
  tibble::as_tibble(as.data.frame(x))
}

first_existing_col <- function(data, candidates) {
  found <- intersect(candidates, names(data))
  if (length(found) == 0) NULL else found[[1]]
}

is_truthy_flag <- function(x) {
  if (is.logical(x)) return(!is.na(x) & x)
  vals <- tolower(trimws(as.character(x)))
  !is.na(vals) & vals %in% c("true", "t", "yes", "y", "1", "+")
}

has_modification <- function(x) {
  vals <- tolower(trimws(as.character(x)))
  !is.na(vals) & nzchar(vals) &
    !vals %in% c("none", "unmodified", "no modifications", "na")
}

add_psm_feature_id <- function(data) {
  data |>
    mutate(
      .qc_feature_id = paste(
        .data$`Master Protein Accessions`,
        .data$`Annotated Sequence`,
        .data$Charge,
        sep = " | "
      )
    )
}

coerce_log2_intensity <- function(x, value_col) {
  vals <- suppressWarnings(as.numeric(x))
  finite_vals <- vals[is.finite(vals)]
  if (length(finite_vals) == 0) return(vals)

  looks_linear <- grepl("^Intensity$|Quan Value", value_col, ignore.case = TRUE) ||
    stats::median(finite_vals, na.rm = TRUE) > 100

  if (looks_linear) {
    log2(pmax(vals, .Machine$double.xmin))
  } else {
    vals
  }
}

impute_matrix_by_row_mean <- function(mat) {
  if (nrow(mat) == 0 || ncol(mat) == 0) return(mat)

  keep <- rowSums(is.finite(mat)) >= 2
  mat <- mat[keep, , drop = FALSE]
  if (nrow(mat) == 0) return(mat)

  row_means <- rowMeans(mat, na.rm = TRUE)
  for (i in seq_len(nrow(mat))) {
    mat[i, !is.finite(mat[i, ])] <- row_means[[i]]
  }

  row_vars <- apply(mat, 1, stats::var, na.rm = TRUE)
  mat[is.finite(row_vars) & row_vars > 0, , drop = FALSE]
}

make_feature_run_matrix <- function(data, feature_col, run_col, value_col,
                                    max_features = params$qc_max_features) {
  matrix_data <- data |>
    mutate(
      .qc_feature = as.character(.data[[feature_col]]),
      .qc_run = as.character(.data[[run_col]]),
      .qc_log2 = coerce_log2_intensity(.data[[value_col]], value_col)
    ) |>
    filter(!is.na(.data$.qc_feature), nzchar(.data$.qc_feature),
           !is.na(.data$.qc_run), nzchar(.data$.qc_run),
           is.finite(.data$.qc_log2)) |>
    group_by(.data$.qc_feature, .data$.qc_run) |>
    summarise(log2Intensity = mean(.data$.qc_log2, na.rm = TRUE),
              .groups = "drop")

  if (nrow(matrix_data) == 0) return(matrix(numeric(), nrow = 0, ncol = 0))

  selected_features <- matrix_data |>
    group_by(.data$.qc_feature) |>
    summarise(
      n_runs = n_distinct(.data$.qc_run),
      variance = stats::var(.data$log2Intensity, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(variance = if_else(is.finite(.data$variance), .data$variance, 0)) |>
    arrange(desc(.data$n_runs), desc(.data$variance)) |>
    slice_head(n = max_features) |>
    pull(.data$.qc_feature)

  wide <- matrix_data |>
    filter(.data$.qc_feature %in% selected_features) |>
    tidyr::pivot_wider(
      names_from = ".qc_run",
      values_from = "log2Intensity"
    )

  mat <- wide |>
    tibble::column_to_rownames(".qc_feature") |>
    as.matrix()
  storage.mode(mat) <- "numeric"
  mat
}

plot_correlation_heatmap <- function(feature_matrix, sample_annotation, title) {
  if (ncol(feature_matrix) < 2 || nrow(feature_matrix) < 2) {
    message("Correlation heatmap skipped: not enough quantified features or runs.")
    return(invisible(NULL))
  }

  mat <- impute_matrix_by_row_mean(feature_matrix)
  if (ncol(mat) < 2 || nrow(mat) < 2) {
    message("Correlation heatmap skipped: not enough complete features after imputation.")
    return(invisible(NULL))
  }

  corr <- stats::cor(mat, use = "pairwise.complete.obs")
  corr_df <- as.data.frame(as.table(corr), stringsAsFactors = FALSE)
  names(corr_df) <- c("Run1", "Run2", "Correlation")

  correlation_annotation <- sample_annotation |>
    mutate(
      CorrelationLabel = if_else(
        !is.na(.data$Condition) & nzchar(.data$Condition) &
          !is.na(.data$BioReplicate) & nzchar(as.character(.data$BioReplicate)),
        paste0(.data$Condition, "_", .data$BioReplicate),
        coalesce(.data$FigureLabel, .data$Run)
      )
    )
  label_lookup <- stats::setNames(
    correlation_annotation$CorrelationLabel,
    correlation_annotation$Run
  )
  display_levels <- unname(label_lookup[colnames(mat)])
  display_levels[is.na(display_levels)] <- colnames(mat)[is.na(display_levels)]
  corr_df <- corr_df |>
    mutate(
      Sample1 = coalesce(unname(label_lookup[as.character(.data$Run1)]), as.character(.data$Run1)),
      Sample2 = coalesce(unname(label_lookup[as.character(.data$Run2)]), as.character(.data$Run2)),
      Sample1 = factor(.data$Sample1, levels = unique(display_levels)),
      Sample2 = factor(.data$Sample2, levels = unique(display_levels))
    )

  ggplot(corr_df, aes(x = Sample1, y = Sample2, fill = Correlation)) +
    geom_tile(color = "white", linewidth = 0.2) +
    scale_fill_gradient2(
      low = "#2C7BB6", mid = "white", high = "#D7191C",
      midpoint = 0.9, limits = c(min(corr_df$Correlation, na.rm = TRUE), 1)
    ) +
    labs(x = NULL, y = NULL, title = title, fill = "Pearson r") +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid = element_blank()
    )
}

plot_pca_from_matrix <- function(feature_matrix, sample_annotation, title) {
  if (ncol(feature_matrix) < 3 || nrow(feature_matrix) < 2) {
    message("PCA skipped: at least 3 runs and 2 quantified features are required.")
    return(invisible(NULL))
  }

  mat <- impute_matrix_by_row_mean(feature_matrix)
  if (ncol(mat) < 3 || nrow(mat) < 2) {
    message("PCA skipped: not enough complete, variable features after imputation.")
    return(invisible(NULL))
  }

  pca <- stats::prcomp(t(mat), center = TRUE, scale. = FALSE)
  var_explained <- (pca$sdev^2) / sum(pca$sdev^2)

  pca_df <- as.data.frame(pca$x[, seq_len(min(2, ncol(pca$x))), drop = FALSE]) |>
    tibble::rownames_to_column("Run") |>
    left_join(sample_annotation, by = "Run") |>
    mutate(
      PcaLabel = if_else(
        !is.na(.data$BioReplicate) & nzchar(as.character(.data$BioReplicate)),
        paste0("BR", .data$BioReplicate),
        coalesce(.data$FigureLabel, .data$Run)
      )
    )

  ggplot(pca_df, aes(x = PC1, y = PC2, color = Condition, label = PcaLabel)) +
    geom_point(size = 3, alpha = 0.9) +
    ggrepel::geom_text_repel(size = 3, max.overlaps = 20, show.legend = FALSE) +
    labs(
      title = title,
      x = paste0("PC1 (", scales::percent(var_explained[[1]], accuracy = 0.1), ")"),
      y = paste0("PC2 (", scales::percent(var_explained[[2]], accuracy = 0.1), ")")
    ) +
    theme_bw(base_size = 12)
}

make_protein_level_matrix <- function(processed_data, max_features = params$qc_max_features) {
  protein_level <- as_plain_tibble(processed_data[["ProteinLevelData"]])
  protein_col <- first_existing_col(protein_level, c("Protein", "PROTEIN", "ProteinName"))
  run_col <- first_existing_col(protein_level, c("Run", "RUN", "originalRUN", "SUBJECT", "BioReplicate"))
  value_col <- first_existing_col(protein_level, c("LogIntensities", "log2Intensity", "ABUNDANCE", "Intensity"))

  if (is.null(protein_col) || is.null(run_col) || is.null(value_col)) {
    message("Protein-level matrix skipped: required MSstats columns were not found.")
    return(matrix(numeric(), nrow = 0, ncol = 0))
  }

  make_feature_run_matrix(protein_level, protein_col, run_col, value_col, max_features)
}

extract_distribution_data <- function(data, stage_label) {
  tbl <- as_plain_tibble(data)
  run_col <- first_existing_col(tbl, c("Run", "RUN", "originalRUN", "SpectrumFile"))
  condition_col <- first_existing_col(tbl, c("Condition", "GROUP", "GROUP_ORIGINAL", "Condition_ORIGINAL"))
  value_col <- first_existing_col(tbl, c("LogIntensities", "log2Intensity", "ABUNDANCE", "Intensity", "Quan Value"))

  if (is.null(run_col) || is.null(value_col)) {
    return(tibble(Stage = character(), Run = character(),
                  Condition = character(), log2Intensity = numeric()))
  }

  tbl |>
    transmute(
      Stage = stage_label,
      Run = as.character(.data[[run_col]]),
      Condition = if (!is.null(condition_col)) as.character(.data[[condition_col]]) else NA_character_,
      log2Intensity = coerce_log2_intensity(.data[[value_col]], value_col)
    ) |>
    filter(is.finite(.data$log2Intensity))
}

extract_normalized_distribution_data <- function(
  data,
  normalization_method = "equalizeMedians",
  stage_label = "After normalization"
) {
  tbl <- as_plain_tibble(data)
  run_col <- first_existing_col(tbl, c("Run", "RUN", "originalRUN", "SpectrumFile"))
  condition_col <- first_existing_col(tbl, c("Condition", "GROUP", "GROUP_ORIGINAL", "Condition_ORIGINAL"))
  value_col <- first_existing_col(tbl, c("LogIntensities", "log2Intensity", "ABUNDANCE", "Intensity", "Quan Value"))
  fraction_col <- first_existing_col(tbl, c("Fraction", "FRACTION"))
  label_col <- first_existing_col(tbl, c("IsotopeLabelType", "LABEL"))

  empty_result <- tibble(
    Stage = character(), Run = character(),
    Condition = character(), log2Intensity = numeric()
  )
  if (is.null(run_col) || is.null(value_col)) return(empty_result)

  distribution <- tbl |>
    transmute(
      Run = as.character(.data[[run_col]]),
      Condition = if (!is.null(condition_col)) as.character(.data[[condition_col]]) else NA_character_,
      Fraction = if (!is.null(fraction_col)) as.character(.data[[fraction_col]]) else "1",
      Label = if (!is.null(label_col)) as.character(.data[[label_col]]) else "L",
      log2Intensity = coerce_log2_intensity(.data[[value_col]], value_col)
    ) |>
    filter(is.finite(.data$log2Intensity))

  method <- toupper(as.character(normalization_method))
  if (method %in% c("NONE", "FALSE")) {
    return(distribution |>
      transmute(Stage = stage_label, .data$Run, .data$Condition, .data$log2Intensity))
  }
  if (method != "EQUALIZEMEDIANS") {
    stop(
      "Normalization diagnostic currently supports equalizeMedians or no normalization; requested: ",
      normalization_method,
      call. = FALSE
    )
  }

  label_values <- unique(distribution$Label[!is.na(distribution$Label)])
  reference_label <- if (length(label_values) == 1L) label_values[[1]] else "H"
  run_medians <- distribution |>
    group_by(.data$Run, .data$Fraction) |>
    summarise(
      RunMedian = median(
        .data$log2Intensity[.data$Label == reference_label],
        na.rm = TRUE
      ),
      .groups = "drop"
    )

  distribution |>
    left_join(run_medians, by = c("Run", "Fraction")) |>
    group_by(.data$Fraction) |>
    mutate(FractionMedian = median(.data$RunMedian, na.rm = TRUE)) |>
    ungroup() |>
    transmute(
      Stage = stage_label,
      .data$Run,
      .data$Condition,
      log2Intensity = .data$log2Intensity - .data$RunMedian + .data$FractionMedian
    ) |>
    filter(is.finite(.data$log2Intensity))
}

summarize_flag_columns <- function(data, patterns) {
  tbl <- as_plain_tibble(data)
  flag_cols <- names(tbl)[stringr::str_detect(tolower(names(tbl)), patterns)]
  if (length(flag_cols) == 0) {
    return(tibble(column = character(), value = character(), n = integer()))
  }

  purrr::map_dfr(flag_cols, function(col) {
    tbl |>
      mutate(.value = as.character(.data[[col]]),
             .value = if_else(is.na(.data$.value) | !nzchar(.data$.value),
                              "Missing/blank", .data$.value)) |>
      count(value = .data$.value, name = "n") |>
      mutate(column = col, .before = 1) |>
      arrange(.data$column, desc(.data$n))
  })
}
```

## Sample Annotation

```{r sample-annotation}
annotation |>
  mutate(.bioreplicate_order = suppressWarnings(as.numeric(.data$BioReplicate))) |>
  arrange(
    .data$Condition,
    is.na(.data$.bioreplicate_order),
    .data$.bioreplicate_order,
    .data$BioReplicate,
    .data$`File Name`
  ) |>
  select(`File Name`, Condition, BioReplicate) |>
  knitr::kable(caption = "Sample annotation used in the MSstats analysis.")
```

## Raw Data Overview

```{r raw-summary}
psm_summary <- tibble(
  metric = c(
    "PSM rows",
    "Runs",
    "Conditions",
    "Protein accessions",
    "Annotated peptide sequences",
    "Non-missing quant values"
  ),
  value = c(
    nrow(psm),
    dplyr::n_distinct(psm$SpectrumFile),
    dplyr::n_distinct(annotation$Condition),
    dplyr::n_distinct(psm$`Master Protein Accessions`, na.rm = TRUE),
    dplyr::n_distinct(psm$`Annotated Sequence`, na.rm = TRUE),
    sum(!is.na(psm$`Quan Value`))
  )
)

knitr::kable(psm_summary, caption = "Summary of the raw PSM dataset.")
```

```{r annotation-summary}
annotation |>
  count(Condition, name = "runs") |>
  arrange(Condition) |>
  knitr::kable(caption = "Number of experimental runs by condition.")
```

```{r raw-psm-counts, fig.cap="PSM counts for each sample before filtering, colored by experimental condition."}
psm |>
  count(SpectrumFile, name = "psms") |>
  left_join(annotation, by = c("SpectrumFile" = "Run")) |>
  ggplot(aes(x = reorder(FigureLabel, psms), y = psms, fill = Condition)) +
  geom_col() +
  coord_flip() +
  labs(x = "Sample", y = "PSM rows", title = "PSM rows per sample") +
  theme_bw()
```

## Raw Data QC

These checks look at the unfiltered PSM export. They are meant to catch run-level
dropout, low-intensity samples, contaminant burden, shared peptide burden, and
unexpected charge or modification patterns before any MSstats modeling.

```{r raw-missingness-zero-summary}
raw_intensity_qc <- psm |>
  left_join(annotation, by = c("SpectrumFile" = "Run")) |>
  group_by(.data$SpectrumFile, .data$FigureLabel, .data$Condition) |>
  summarise(
    psm_rows = n(),
    missing_intensity = sum(is.na(.data$`Quan Value`)),
    non_positive_intensity = sum(!is.na(.data$`Quan Value`) & .data$`Quan Value` <= 0),
    positive_intensity = sum(!is.na(.data$`Quan Value`) & .data$`Quan Value` > 0),
    missing_or_non_positive_percent =
      100 * (.data$missing_intensity + .data$non_positive_intensity) / .data$psm_rows,
    .groups = "drop"
  ) |>
  arrange(.data$Condition, .data$SpectrumFile)

knitr::kable(
  raw_intensity_qc,
  digits = 2,
  caption = "Raw intensity completeness and non-positive values by sample."
)
```

```{r raw-missingness-zero-plot, fig.height=6, fig.cap="Raw PSM intensity-status counts per sample, separating positive, non-positive, and missing quantitative values."}
raw_intensity_qc |>
  tidyr::pivot_longer(
    cols = c("missing_intensity", "non_positive_intensity", "positive_intensity"),
    names_to = "status",
    values_to = "rows"
  ) |>
  mutate(
    status = factor(
      .data$status,
      levels = c("positive_intensity", "non_positive_intensity", "missing_intensity"),
      labels = c("Positive", "Non-positive", "Missing")
    )
  ) |>
  ggplot(aes(x = reorder(.data$FigureLabel, .data$rows), y = .data$rows, fill = .data$status)) +
  geom_col() +
  coord_flip() +
  labs(x = "Sample", y = "PSM rows", fill = "Intensity status",
       title = "Raw PSM intensity status per sample") +
  theme_bw()
```

```{r raw-intensity-distribution, fig.height=7, fig.cap="Distribution of raw log2 PSM intensities for each sample, colored by condition."}
psm |>
  filter(!is.na(.data$`Quan Value`), .data$`Quan Value` > 0) |>
  left_join(annotation, by = c("SpectrumFile" = "Run")) |>
  mutate(log2QuanValue = log2(.data$`Quan Value`)) |>
  ggplot(aes(x = reorder(.data$FigureLabel, .data$log2QuanValue, median),
             y = .data$log2QuanValue, fill = .data$Condition)) +
  geom_boxplot(outlier.alpha = 0.15, outlier.size = 0.4) +
  coord_flip() +
  labs(x = "Sample", y = "log2(Quan Value)",
       title = "Raw intensity distribution per sample") +
  theme_bw()
```

```{r raw-intensity-density, fig.height=6, fig.cap="Kernel-density curves of raw log2 PSM intensities for each sample, colored by condition."}
psm |>
  filter(!is.na(.data$`Quan Value`), .data$`Quan Value` > 0) |>
  left_join(annotation, by = c("SpectrumFile" = "Run")) |>
  mutate(log2QuanValue = log2(.data$`Quan Value`)) |>
  ggplot(aes(x = .data$log2QuanValue, color = .data$Condition, group = .data$SpectrumFile)) +
  geom_density(alpha = 0.45, linewidth = 0.5) +
  labs(x = "log2(Quan Value)", y = "Density",
       title = "Raw intensity density per run") +
  theme_bw()
```

```{r raw-contaminant-shared-single-summary}
contaminant_rows <- if ("Contaminant" %in% names(psm)) {
  sum(is_truthy_flag(psm$Contaminant))
} else {
  NA_integer_
}

shared_peptide_rows <- if ("# Proteins" %in% names(psm)) {
  sum(suppressWarnings(as.numeric(psm$`# Proteins`)) > 1, na.rm = TRUE)
} else {
  NA_integer_
}

single_peptide_proteins <- psm |>
  filter(!is.na(.data$`Master Protein Accessions`),
         !is.na(.data$`Annotated Sequence`)) |>
  distinct(.data$`Master Protein Accessions`, .data$`Annotated Sequence`) |>
  count(.data$`Master Protein Accessions`, name = "unique_peptides") |>
  summarise(n = sum(.data$unique_peptides == 1)) |>
  pull(.data$n)

tibble(
  metric = c(
    "Contaminant PSM rows",
    "Shared peptide PSM rows (# Proteins > 1)",
    "Proteins represented by one unique peptide"
  ),
  value = c(contaminant_rows, shared_peptide_rows, single_peptide_proteins)
) |>
  knitr::kable(
    caption = "Contaminant, shared-peptide, and single-peptide-protein summary."
  )
```

```{r raw-charge-distribution, fig.height=5, fig.cap="Raw PSM counts by precursor charge state."}
if ("Charge" %in% names(psm)) {
  psm |>
    mutate(Charge = as.character(.data$Charge)) |>
    count(.data$Charge, name = "psm_rows") |>
    ggplot(aes(x = .data$Charge, y = .data$psm_rows)) +
    geom_col(fill = "#4C72B0") +
    labs(x = "Charge", y = "PSM rows", title = "Charge-state distribution") +
    theme_bw()
} else {
  message("Charge-state plot skipped: Charge column was not found.")
}
```

```{r raw-modification-distribution, fig.height=5, fig.cap="Raw PSM counts for modified and unmodified peptide-spectrum matches."}
if ("Modifications" %in% names(psm)) {
  psm |>
    mutate(
      modification_status = if_else(has_modification(.data$Modifications),
                                    "Modified", "Unmodified")
    ) |>
    count(.data$modification_status, name = "psm_rows") |>
    ggplot(aes(x = .data$modification_status, y = .data$psm_rows,
               fill = .data$modification_status)) +
    geom_col(show.legend = FALSE) +
    labs(x = NULL, y = "PSM rows", title = "Modified vs unmodified PSM rows") +
    theme_bw()
} else {
  message("Modification plot skipped: Modifications column was not found.")
}
```

## Preprocessing for MSstats

The PSM table is filtered before conversion. By default this report keeps high-confidence PSMs, removes rows flagged as contaminants when that column is present, and removes non-positive or missing intensities.

```{r preprocess-for-msstats}
psm_filtered <- psm |>
  filter(!is.na(`Quan Value`), `Quan Value` > 0)

if (!is.null(params$filter_confidence) && "Confidence" %in% names(psm_filtered)) {
  psm_filtered <- psm_filtered |>
    filter(.data$Confidence == params$filter_confidence)
}

if (isTRUE(params$remove_contaminants) && "Contaminant" %in% names(psm_filtered)) {
  psm_filtered <- psm_filtered |>
    filter(is.na(.data$Contaminant) | .data$Contaminant == FALSE)
}

preprocess_summary <- tibble(
  step = c("Raw PSM rows", "Rows after filtering"),
  rows = c(nrow(psm), nrow(psm_filtered))
)

knitr::kable(
  preprocess_summary,
  caption = "PSM row counts before and after MSstats input filtering."
)
```

## Filtered Input QC

These summaries describe the data that will actually be submitted to
`MSstats::PDtoMSstatsFormat()`.

```{r filtered-feature-counts}
psm_filtered_features <- add_psm_feature_id(psm_filtered)

filtered_run_qc <- psm_filtered_features |>
  left_join(annotation, by = c("SpectrumFile" = "Run")) |>
  group_by(.data$SpectrumFile, .data$FigureLabel, .data$Condition) |>
  summarise(
    psm_rows = n(),
    annotated_sequences = n_distinct(.data$`Annotated Sequence`, na.rm = TRUE),
    peptide_charge_features = n_distinct(.data$.qc_feature_id, na.rm = TRUE),
    protein_accessions = n_distinct(.data$`Master Protein Accessions`, na.rm = TRUE),
    median_log2_intensity = median(log2(.data$`Quan Value`), na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(.data$Condition, .data$SpectrumFile)

knitr::kable(
  filtered_run_qc,
  digits = 2,
  caption = "Filtered MSstats input features and intensities by sample."
)
```

```{r filtered-feature-counts-plot, fig.height=7, fig.cap="Filtered PSM, annotated-sequence, peptide-charge-feature, and protein-accession counts for each sample."}
filtered_run_qc |>
  select(SpectrumFile, FigureLabel, Condition, psm_rows, annotated_sequences,
         peptide_charge_features, protein_accessions) |>
  tidyr::pivot_longer(
    cols = c("psm_rows", "annotated_sequences",
             "peptide_charge_features", "protein_accessions"),
    names_to = "metric",
    values_to = "count"
  ) |>
  mutate(
    metric = factor(
      .data$metric,
      levels = c("psm_rows", "annotated_sequences",
                 "peptide_charge_features", "protein_accessions"),
      labels = c("PSM rows", "Annotated sequences",
                 "Peptide-charge features", "Protein accessions")
    )
  ) |>
  ggplot(aes(x = reorder(.data$FigureLabel, .data$count),
             y = .data$count, fill = .data$Condition)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ metric, scales = "free_x") +
  labs(x = "Sample", y = "Count",
       title = "Filtered PSM, peptide, feature, and protein counts per sample") +
  theme_bw()
```

```{r filtered-run-correlation, fig.height=8, fig.cap="Pearson correlations among filtered peptide-charge feature profiles; samples are labeled by condition and biological replicate."}
filtered_feature_matrix <- make_feature_run_matrix(
  psm_filtered_features,
  feature_col = ".qc_feature_id",
  run_col = "SpectrumFile",
  value_col = "Quan Value"
)

plot_correlation_heatmap(
  filtered_feature_matrix,
  annotation,
  title = "Filtered PSM feature correlation between runs"
)
```

```{r helper-functions}
convert_pd_to_msstats <- function(psm_table, annotation_table) {
  MSstats::PDtoMSstatsFormat(
    input = psm_table,
    annotation = annotation_table,
    useUniquePeptide = params$remove_shared_peptides,
    summaryforMultipleRows = max,
    removeFewMeasurements = params$remove_few_measurements,
    removeProtein_with1Peptide = params$remove_single_peptide_proteins,
    which.quantification = "Quan Value",
    which.proteinid = "Master Protein Accessions",
    which.sequence = "Annotated Sequence",
    use_log_file = FALSE,
    verbose = FALSE
  )
}

process_msstats <- function(msstats_input) {
  MSstats::dataProcess(
    raw = msstats_input,
    logTrans = 2,
    normalization = params$normalization,
    summaryMethod = params$summary_method,
    MBimpute = params$mb_impute,
    use_log_file = FALSE,
    verbose = FALSE,
    numberOfCores = params$number_of_cores
  )
}

make_contrast_matrix <- function(groups, contrast_table) {
  missing_groups <- setdiff(unique(c(contrast_table$numerator, contrast_table$denominator)), groups)
  if (length(missing_groups) > 0) {
    stop("Contrast group(s) not present in processed data: ", paste(missing_groups, collapse = ", "), call. = FALSE)
  }

  contrast_matrix <- matrix(
    0,
    nrow = nrow(contrast_table),
    ncol = length(groups),
    dimnames = list(contrast_table$label, groups)
  )
  for (i in seq_len(nrow(contrast_table))) {
    contrast_matrix[i, contrast_table$numerator[i]] <- 1
    contrast_matrix[i, contrast_table$denominator[i]] <- -1
  }
  contrast_matrix
}

make_weighted_contrast_matrix <- function(groups, label, weights) {
  missing_groups <- setdiff(names(weights), groups)
  if (length(missing_groups) > 0) {
    stop("Contrast group(s) not present in processed data: ", paste(missing_groups, collapse = ", "), call. = FALSE)
  }

  contrast_matrix <- matrix(
    0,
    nrow = 1,
    ncol = length(groups),
    dimnames = list(label, groups)
  )
  contrast_matrix[label, names(weights)] <- weights
  contrast_matrix
}

run_group_comparison <- function(processed_data, contrast_table) {
  groups <- processed_groups(processed_data)
  contrast_matrix <- make_contrast_matrix(groups, contrast_table)

  run_group_comparison_matrix(processed_data, contrast_matrix)
}

run_group_comparison_matrix <- function(processed_data, contrast_matrix) {
  comparison <- MSstats::groupComparison(
    contrast.matrix = contrast_matrix,
    data = processed_data,
    use_log_file = FALSE,
    verbose = FALSE,
    numberOfCores = params$number_of_cores
  )

  list(
    contrast_matrix = contrast_matrix,
    comparison = comparison,
    results = as_plain_tibble(comparison[["ComparisonResult"]])
  )
}

choose_break_interval <- function(values, side, quantile_prob, min_ratio) {
  values <- values[is.finite(values)]
  if (length(values) == 0) return(NULL)

  abs_values <- abs(values)
  core_limit <- max(
    stats::quantile(abs_values, quantile_prob, na.rm = TRUE, names = FALSE),
    params$volcano_fc_cutoff * 1.5,
    2
  )
  max_abs <- max(abs_values, na.rm = TRUE)
  if (!is.finite(max_abs) || max_abs < core_limit * min_ratio) return(NULL)

  if (side == "positive") {
    extreme_edge <- min(values[values > core_limit], na.rm = TRUE)
    if (!is.finite(extreme_edge) || extreme_edge <= core_limit) return(NULL)
    c(core_limit, extreme_edge * 0.9)
  } else {
    extreme_edge <- max(values[values < -core_limit], na.rm = TRUE)
    if (!is.finite(extreme_edge) || extreme_edge >= -core_limit) return(NULL)
    c(extreme_edge * 0.9, -core_limit)
  }
}

choose_y_break_interval <- function(neg_log10_p, quantile_prob, min_ratio) {
  neg_log10_p <- neg_log10_p[is.finite(neg_log10_p)]
  if (length(neg_log10_p) == 0) return(NULL)

  core_limit <- max(
    stats::quantile(neg_log10_p, quantile_prob, na.rm = TRUE, names = FALSE),
    -log10(params$volcano_p_cutoff) * 1.5,
    5
  )
  max_y <- max(neg_log10_p, na.rm = TRUE)
  if (!is.finite(max_y) || max_y < core_limit * min_ratio) return(NULL)

  extreme_edge <- min(neg_log10_p[neg_log10_p > core_limit], na.rm = TRUE)
  if (!is.finite(extreme_edge) || extreme_edge <= core_limit) return(NULL)

  c(core_limit, extreme_edge * 0.9)
}

apply_split_axes <- function(plot, plot_data, fc_col, p_col) {
  split_scales <- params$volcano_split_axis_scales
  split_quantile <- params$volcano_split_axis_quantile
  split_min_ratio <- params$volcano_split_axis_min_ratio

  x_values <- plot_data[[fc_col]]
  x_neg_break <- choose_break_interval(x_values, "negative", split_quantile, split_min_ratio)
  x_pos_break <- choose_break_interval(x_values, "positive", split_quantile, split_min_ratio)
  y_break <- choose_y_break_interval(-log10(plot_data[[p_col]]), split_quantile, split_min_ratio)

  if (!is.null(x_neg_break) && diff(x_neg_break) > 0) {
    plot <- plot + ggbreak::scale_x_break(x_neg_break, scales = split_scales)
  }
  if (!is.null(x_pos_break) && diff(x_pos_break) > 0) {
    plot <- plot + ggbreak::scale_x_break(x_pos_break, scales = split_scales)
  }
  if (!is.null(y_break) && diff(y_break) > 0) {
    plot <- plot + ggbreak::scale_y_break(y_break, scales = split_scales)
  }

  plot
}

plot_volcano <- function(results, comparison_label, p_col = "adj.pvalue", fc_col = "log2FC") {
  plot_data <- results |>
    filter(.data$Label == comparison_label) |>
    mutate(
      .volcano_p = pmax(.data[[p_col]], .Machine$double.xmin)
    ) |>
    filter(is.finite(.data[[fc_col]]), is.finite(.data$.volcano_p))

  plot_data <- plot_data |>
    mutate(
      GeneSymbol = map_uniprot_to_gene_symbol(.data$Protein),
      .volcano_label = if_else(
        !is.na(.data$GeneSymbol) & nzchar(.data$GeneSymbol),
        .data$GeneSymbol,
        ""
      )
    )

  top_labels <- plot_data |>
    filter(
      nzchar(.data$.volcano_label),
      .data$.volcano_p < params$volcano_p_cutoff
    ) |>
    arrange(.data$.volcano_p, desc(abs(.data[[fc_col]]))) |>
    slice_head(n = params$volcano_top_labels) |>
    pull(.data$.volcano_label)

  if (length(top_labels) == 0) {
    top_labels <- plot_data |>
      filter(nzchar(.data$.volcano_label)) |>
      arrange(.data$.volcano_p, desc(abs(.data[[fc_col]]))) |>
      slice_head(n = params$volcano_top_labels) |>
      pull(.data$.volcano_label)
  }

  volcano_plot <- EnhancedVolcano::EnhancedVolcano(
    plot_data,
    lab = plot_data$.volcano_label,
    x = fc_col,
    y = ".volcano_p",
    selectLab = top_labels,
    pCutoff = params$volcano_p_cutoff,
    FCcutoff = params$volcano_fc_cutoff,
    title = gsub(
      "_", " ",
      sub("_vs_", "/", comparison_label, fixed = TRUE),
      fixed = TRUE
    ),
    subtitle = "MSstats protein-level comparison",
    caption = paste0(
      "Adjusted p < ", params$volcano_p_cutoff,
      "; |log2FC| > ", params$volcano_fc_cutoff
    ),
    xlab = "Log2 fold change",
    ylab = "-Log10 adjusted p-value",
    pointSize = 2.2,
    labSize = 4.0,
    col = c("grey78", "grey45", "#4C72B0", "#C44E52"),
    colAlpha = 0.82,
    legendLabels = c(
      "Not significant",
      "Meets absolute log2 fold-change threshold only",
      "Meets adjusted p-value threshold only",
      "Meets both adjusted p-value and fold-change thresholds"
    ),
    legendPosition = "bottom",
    legendLabSize = 9,
    legendIconSize = 3.5,
    drawConnectors = TRUE,
    widthConnectors = 0.35,
    colConnectors = "grey35"
  ) +
    theme_classic(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 16, margin = margin(b = 5)),
      plot.subtitle = element_text(size = 11, color = "grey30", margin = margin(b = 8)),
      plot.caption = element_text(size = 9, color = "grey30", margin = margin(t = 8)),
      axis.title = element_text(face = "bold", size = 12),
      axis.text = element_text(size = 10, color = "black"),
      legend.title = element_blank(),
      legend.position = "bottom",
      legend.box = "vertical",
      legend.text = element_text(size = 9),
      legend.margin = margin(t = 8),
      plot.margin = margin(12, 18, 12, 12)
    ) +
    guides(
      colour = guide_legend(nrow = 2, byrow = TRUE),
      shape = guide_legend(nrow = 2, byrow = TRUE)
    )

  static_plot <- apply_split_axes(volcano_plot, plot_data, fc_col, ".volcano_p")
  static_file <- file.path(
    figure_dir,
    paste0(params$request, "-", tolower(comparison_label), "-volcano.pdf")
  )
  ggplot2::ggsave(
    static_file,
    plot = static_plot,
    width = 12,
    height = 10,
    units = "in",
    bg = "white"
  )

  static_plot
}

create_dt <- function(data, download_basename, page_length = 25, filter = "top") {
  DT::datatable(
    data,
    rownames = FALSE,
    filter = filter,
    extensions = c("Buttons", "Scroller", "SearchBuilder"),
    width = "100%",
    options = list(
      dom = "QBlrtip",
      scrollX = TRUE,
      scrollY = 650,
      scrollCollapse = TRUE,
      scroller = TRUE,
      deferRender = TRUE,
      pageLength = page_length,
      lengthMenu = list(
        c(10, 25, 50, 100, -1),
        c("10", "25", "50", "100", "All")
      ),
      buttons = list(
        list(extend = "copy"),
        list(extend = "csv", filename = download_basename),
        list(extend = "excel", filename = download_basename)
      ),
      columnDefs = list(list(width = "42%", targets = 5))
    )
  )
}

format_p_values <- function(data) {
  p_cols <- intersect(
    c("adj.pvalue", "pvalue", "p_val_adj", "p_val"),
    names(data)
  )
  data |>
    dplyr::mutate(dplyr::across(
      dplyr::all_of(p_cols),
      ~ ifelse(is.na(.x), NA_character_, formatC(.x, format = "e", digits = 2))
    ))
}

prepare_de_table <- function(results, comparison_label) {
  result_data <- add_power_metrics(
    results,
    alpha = params$power_alpha,
    target_power = params$power_target,
    effect_log2fc = params$power_effect_log2fc
  ) |>
    as.data.frame()

  for (column in c(
    "issue", "MissingPercentage", "ImputationPercentage",
    "target_log2fc_power", "detectable_log2fc_at_target_power",
    "observed_effect_power"
  )) {
    if (!column %in% names(result_data)) result_data[[column]] <- NA
  }

  result_data |>
    as_tibble() |>
    filter(
      .data$Label == comparison_label,
      is.finite(.data$log2FC),
      is.finite(.data$pvalue),
      is.finite(.data$adj.pvalue),
      .data$adj.pvalue < params$volcano_p_cutoff,
      abs(.data$log2FC) > params$volcano_fc_cutoff
    ) |>
    mutate(
      Accession = .data$Protein,
      Gene = map_uniprot_to_gene_symbol(.data$Protein),
      Note = purrr::pmap_chr(
        list(
          .data$issue, .data$MissingPercentage, .data$ImputationPercentage,
          .data$Gene, .data$target_log2fc_power,
          .data$detectable_log2fc_at_target_power, .data$observed_effect_power
        ),
        function(issue, missing, imputation, gene, target_power, detectable_effect,
                 observed_power) {
          flags <- character()
          issue_text <- trimws(as.character(issue))
          if (!is.na(issue_text) && nzchar(issue_text) &&
              !tolower(issue_text) %in% c("na", "none", "false", "0")) {
            flags <- c(flags, paste0("MSstats flag: ", issue_text))
          }
          missing_value <- suppressWarnings(as.numeric(missing))
          if (is.finite(missing_value) && missing_value > 0) {
            flags <- c(flags, paste0("Missing measurements: ", round(missing_value, 1), "%"))
          }
          imputation_value <- suppressWarnings(as.numeric(imputation))
          if (is.finite(imputation_value) && imputation_value > 0) {
            flags <- c(flags, paste0("Imputed measurements: ", round(imputation_value, 1), "%"))
          }
          if (is.na(gene) || !nzchar(gene)) {
            flags <- c(flags, "Gene symbol unavailable")
          }
          target_power_value <- suppressWarnings(as.numeric(target_power))
          detectable_effect_value <- suppressWarnings(as.numeric(detectable_effect))
          if (is.finite(target_power_value) && target_power_value < params$power_target) {
            power_note <- paste0(
              "Power caution: ", scales::percent(target_power_value, accuracy = 1),
              " power for target |log2FC| = ", abs(params$power_effect_log2fc),
              " (below ", scales::percent(params$power_target, accuracy = 1), ")"
            )
            if (is.finite(detectable_effect_value)) {
              power_note <- paste0(
                power_note, "; estimated minimum |log2FC| = ",
                round(detectable_effect_value, 2)
              )
            }
            flags <- c(flags, power_note)
          }
          observed_power_value <- suppressWarnings(as.numeric(observed_power))
          if (is.finite(observed_power_value) && observed_power_value < params$power_target) {
            flags <- c(
              flags,
              paste0(
                "Observed-effect detectability is ",
                scales::percent(observed_power_value, accuracy = 1),
                "; interpret cautiously (not independent evidence)"
              )
            )
          }
          if (length(flags) == 0) "No additional flags." else paste(flags, collapse = "; ")
        }
      )
    ) |>
    arrange(.data$adj.pvalue, desc(abs(.data$log2FC))) |>
    select(Accession, Gene, log2FC, pvalue, adj.pvalue, Note) |>
    format_p_values()
}

as_plain_tibble <- function(x) {
  tibble::as_tibble(as.data.frame(x))
}

condition_column <- function(data) {
  candidates <- c("GROUP_ORIGINAL", "GROUP", "Condition", "Condition_ORIGINAL")
  found <- intersect(candidates, names(data))
  if (length(found) == 0) {
    stop(
      "Could not find a condition/group column. Available columns: ",
      paste(names(data), collapse = ", "),
      call. = FALSE
    )
  }
  found[1]
}

processed_groups <- function(processed_data) {
  protein_level <- as_plain_tibble(processed_data[["ProteinLevelData"]])
  group_col <- condition_column(protein_level)
  sort(unique(as.character(protein_level[[group_col]])))
}

protein_level_group_summary <- function(processed_data) {
  protein_level <- as_plain_tibble(processed_data[["ProteinLevelData"]])
  group_col <- condition_column(protein_level)

  protein_level |>
    count(Condition = .data[[group_col]], name = "protein_summaries") |>
    arrange(.data$Condition)
}

write_results <- function(results, file_name) {
  output_file <- file.path(output_dir, file_name)
  readr::write_csv(results, output_file)
  output_file
}

quietly_run <- function(expr) {
  result <- NULL
  invisible(utils::capture.output(
    invisible(utils::capture.output(result <- force(expr), type = "message")),
    type = "output"
  ))
  result
}
```

## MSstats Processing QC

These diagnostics evaluate the converted and processed MSstats data before group
comparisons are interpreted.

```{r convert-three-condition}
msstats_input <- quietly_run(convert_pd_to_msstats(psm_filtered, annotation))

as_plain_tibble(msstats_input) |>
  count(Condition, Run, name = "features") |>
  arrange(Condition, Run) |>
  head(20) |>
  knitr::kable(
    caption = "Converted MSstats feature counts by condition and run (first 20 rows)."
  )
```

```{r process-three-condition}
processed <- quietly_run(process_msstats(msstats_input))

protein_level_group_summary(processed) |>
  knitr::kable(
    caption = "MSstats protein-level measurement summary by experimental condition."
  )
```

```{r normalization-diagnostic, fig.height=8, fig.cap="Sample intensity distributions before and after MSstats normalization."}
normalization_distribution <- bind_rows(
  extract_distribution_data(msstats_input, "Before normalization"),
  extract_normalized_distribution_data(msstats_input, params$normalization)
)

if (nrow(normalization_distribution) > 0) {
  normalization_distribution |>
    left_join(
      annotation |> select(.data$Run, .data$FigureLabel),
      by = "Run"
    ) |>
    mutate(FigureLabel = coalesce(.data$FigureLabel, .data$Run)) |>
    ggplot(aes(x = reorder(.data$FigureLabel, .data$log2Intensity, median),
               y = .data$log2Intensity, fill = .data$Stage)) +
    geom_boxplot(outlier.alpha = 0.08, outlier.size = 0.3) +
    coord_flip() +
    facet_wrap(~ Stage, scales = "free_x") +
    labs(x = "Sample", y = "log2 intensity",
         title = "Intensity distributions before and after MSstats normalization") +
    theme_bw()
} else {
  message("Normalization diagnostic skipped: intensity columns were not found.")
}
```

```{r processed-sample-annotation, include=FALSE}
protein_level_tbl <- as_plain_tibble(processed[["ProteinLevelData"]])
protein_run_col <- first_existing_col(
  protein_level_tbl,
  c("Run", "RUN", "originalRUN", "SUBJECT", "BioReplicate")
)
protein_condition_col <- first_existing_col(
  protein_level_tbl,
  c("GROUP_ORIGINAL", "GROUP", "Condition", "Condition_ORIGINAL")
)

processed_sample_annotation <- if (!is.null(protein_run_col)) {
  processed_subjects <- protein_level_tbl |>
    transmute(
      Run = as.character(.data[[protein_run_col]]),
      Condition = if (!is.null(protein_condition_col)) {
        as.character(.data[[protein_condition_col]])
      } else {
        NA_character_
      }
    ) |>
    distinct() |>
    mutate(.run_order = suppressWarnings(as.numeric(.data$Run))) |>
    arrange(.data$Condition, is.na(.data$.run_order), .data$.run_order, .data$Run) |>
    group_by(.data$Condition) |>
    mutate(.condition_index = row_number()) |>
    ungroup()

  annotation_subjects <- annotation |>
    mutate(.bioreplicate_order = suppressWarnings(as.numeric(.data$BioReplicate))) |>
    arrange(.data$Condition, is.na(.data$.bioreplicate_order),
            .data$.bioreplicate_order, .data$BioReplicate) |>
    group_by(.data$Condition) |>
    mutate(.condition_index = row_number()) |>
    ungroup() |>
    select(.data$Condition, .data$.condition_index, .data$SampleName,
           .data$BioReplicate, .data$FigureLabel)

  processed_subjects |>
    left_join(annotation_subjects, by = c("Condition", ".condition_index")) |>
    select(Run, SampleName, BioReplicate, FigureLabel, Condition) |>
    distinct()
} else {
  annotation
}

```

```{r protein-level-correlation, fig.height=8, fig.cap="Pearson correlations among MSstats protein-level abundance profiles; samples are labeled by condition and biological replicate."}
protein_level_matrix <- make_protein_level_matrix(processed)

plot_correlation_heatmap(
  protein_level_matrix,
  processed_sample_annotation,
  title = "MSstats protein-level correlation between samples"
)
```

```{r protein-level-pca, fig.height=7, fig.cap="Principal component analysis of MSstats protein-level abundance profiles, colored by condition and labeled by biological replicate (BR#)."}
plot_pca_from_matrix(
  protein_level_matrix,
  processed_sample_annotation,
  title = "MSstats protein-level PCA"
)
```

```{r protein-level-cv-summary}
protein_col <- first_existing_col(protein_level_tbl, c("Protein", "PROTEIN", "ProteinName"))
condition_col <- first_existing_col(protein_level_tbl, c("GROUP_ORIGINAL", "GROUP", "Condition", "Condition_ORIGINAL"))
value_col <- first_existing_col(protein_level_tbl, c("LogIntensities", "log2Intensity", "ABUNDANCE", "Intensity"))

if (!is.null(protein_col) && !is.null(condition_col) && !is.null(value_col)) {
  protein_cv <- protein_level_tbl |>
    transmute(
      Protein = as.character(.data[[protein_col]]),
      Condition = as.character(.data[[condition_col]]),
      linearIntensity = 2^coerce_log2_intensity(.data[[value_col]], value_col)
    ) |>
    filter(is.finite(.data$linearIntensity), .data$linearIntensity > 0) |>
    group_by(.data$Protein, .data$Condition) |>
    summarise(
      n_samples = n(),
      mean_intensity = mean(.data$linearIntensity, na.rm = TRUE),
      sd_intensity = stats::sd(.data$linearIntensity, na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(cv_percent = 100 * .data$sd_intensity / .data$mean_intensity) |>
    filter(.data$n_samples >= 2, is.finite(.data$cv_percent))

  protein_cv |>
    group_by(.data$Condition) |>
    summarise(
      proteins = n(),
      median_cv_percent = median(.data$cv_percent, na.rm = TRUE),
      p75_cv_percent = stats::quantile(.data$cv_percent, 0.75, na.rm = TRUE),
      p90_cv_percent = stats::quantile(.data$cv_percent, 0.90, na.rm = TRUE),
      .groups = "drop"
    ) |>
    knitr::kable(
      digits = 2,
      caption = "Within-condition protein abundance coefficient-of-variation summary."
    )
} else {
  protein_cv <- tibble()
  message("Protein-level CV skipped: required protein, condition, or intensity columns were not found.")
}
```

```{r protein-level-cv-plot, fig.height=6, fig.cap="Distributions of within-condition protein-level coefficients of variation."}
if (nrow(protein_cv) > 0) {
  ggplot(protein_cv, aes(x = .data$Condition, y = .data$cv_percent,
                         fill = .data$Condition)) +
    geom_violin(trim = TRUE, alpha = 0.65, show.legend = FALSE) +
    geom_boxplot(width = 0.18, outlier.alpha = 0.15, show.legend = FALSE) +
    labs(x = NULL, y = "CV (%)",
         title = "Protein-level coefficient of variation within condition") +
    theme_bw()
}
```

```{r top-variable-proteins}
if (nrow(protein_cv) > 0) {
  protein_cv |>
    arrange(desc(.data$cv_percent)) |>
    mutate(
      `Gene names` = map_uniprot_to_gene_symbol(.data$Protein),
      .after = Protein
    ) |>
    select(Protein, `Gene names`, Condition, n_samples, cv_percent) |>
    slice_head(n = params$qc_top_cv_proteins) |>
    knitr::kable(
      digits = 2,
      caption = "Proteins with the highest within-condition coefficients of variation."
    )
}
```

## MSstats Analysis: Healthy, FC, and LC

This analysis preserves the three conditions in the annotation file and tests FC vs Healthy, LC vs Healthy, and LC vs FC.

```{r compare-three-condition}
three_condition_contrasts <- tibble::tribble(
  ~label, ~numerator, ~denominator,
  "FC_vs_Healthy", "FC", "Healthy",
  "LC_vs_Healthy", "LC", "Healthy",
  "LC_vs_FC", "LC", "FC"
)

three_condition_contrast_matrix <- make_contrast_matrix(
  processed_groups(processed),
  three_condition_contrasts
)

infected_contrast_matrix <- make_weighted_contrast_matrix(
  groups = processed_groups(processed),
  label = "Infected_vs_Healthy",
  weights = c(FC = 0.5, LC = 0.5, Healthy = -1)
)

all_contrast_matrix <- rbind(
  three_condition_contrast_matrix,
  infected_contrast_matrix
)

all_comparison_results <- quietly_run(
  run_group_comparison_matrix(processed, all_contrast_matrix)
)
all_comparison_output <- all_comparison_results$results

saveRDS(
  list(
    msstats_input = msstats_input,
    processed_data = processed[["ProcessedData"]],
    protein_level_data = processed[["ProteinLevelData"]],
    annotation = annotation,
    comparison_results = all_comparison_output
  ),
  file.path(output_dir, "msstats_workbook_export_objects.rds")
)

three_condition_output <- all_comparison_output |>
  filter(.data$Label %in% three_condition_contrasts$label)
infected_output <- all_comparison_output |>
  filter(.data$Label == "Infected_vs_Healthy")

three_condition_file <- write_results(three_condition_output, "msstats_three_condition_group_comparison.csv")
infected_file <- write_results(infected_output, "msstats_infected_vs_healthy_group_comparison.csv")

knitr::kable(
  three_condition_contrast_matrix,
  caption = "Contrast matrix for FC, LC, and Healthy comparisons."
)
```

Volcano plots are publication-quality static `EnhancedVolcano` figures. Labels
use gene symbols, connectors reduce label ambiguity, and the complete
significance legend is written below each plot. Automatic `ggbreak` split axes
keep extreme tails from compressing the main protein cloud. Each figure is also
exported as a vector PDF in the `figures` directory.

::: {.panel-tabset .volcano-tabset}

### FC vs Healthy

```{r fig-volcano-fc-vs-healthy, fig.width=12, fig.height=10, dpi=300, out.width="100%", fig.cap="Protein-level volcano plot for FC versus Healthy. Dashed thresholds show the configured adjusted-p-value and absolute log2-fold-change cutoffs."}
plot_volcano(three_condition_output, "FC_vs_Healthy")
```

#### Differentially expressed proteins

```{r de-table-fc-vs-healthy}
prepare_de_table(three_condition_output, "FC_vs_Healthy") |>
  create_dt(paste0(params$request, "-FC-over-Healthy-DE-proteins"))
```

### LC vs Healthy

```{r fig-volcano-lc-vs-healthy, fig.width=12, fig.height=10, dpi=300, out.width="100%", fig.cap="Protein-level volcano plot for LC versus Healthy. Dashed thresholds show the configured adjusted-p-value and absolute log2-fold-change cutoffs."}
plot_volcano(three_condition_output, "LC_vs_Healthy")
```

#### Differentially expressed proteins

```{r de-table-lc-vs-healthy}
prepare_de_table(three_condition_output, "LC_vs_Healthy") |>
  create_dt(paste0(params$request, "-LC-over-Healthy-DE-proteins"))
```

### LC vs FC

```{r fig-volcano-lc-vs-fc, fig.width=12, fig.height=10, dpi=300, out.width="100%", fig.cap="Protein-level volcano plot for LC versus FC. Dashed thresholds show the configured adjusted-p-value and absolute log2-fold-change cutoffs."}
plot_volcano(three_condition_output, "LC_vs_FC")
```

#### Differentially expressed proteins

```{r de-table-lc-vs-fc}
prepare_de_table(three_condition_output, "LC_vs_FC") |>
  create_dt(paste0(params$request, "-LC-over-FC-DE-proteins"))
```

:::

## MSstats Analysis: Infected vs Healthy

This analysis tests the average infected effect from the three-condition model: `(FC + LC) / 2 - Healthy`. It reuses the processed MSstats model above instead of collapsing FC and LC before conversion.

```{r infected-annotation}
annotation |>
  mutate(InfectionStatus = if_else(Condition == "Healthy", "Healthy", "Infected")) |>
  count(InfectionStatus, Condition, name = "runs") |>
  arrange(InfectionStatus, Condition) |>
  knitr::kable(
    caption = "Runs contributing to the Infected-versus-Healthy comparison."
  )
```

```{r compare-infected}
knitr::kable(
  infected_contrast_matrix,
  caption = "Contrast matrix for the Infected-versus-Healthy comparison."
)
```

The infected contrast uses the same publication-quality static volcano styling
as the three-condition comparisons.

```{r fig-infected-volcano, fig.width=12, fig.height=10, dpi=300, out.width="100%", fig.cap="Protein-level volcano plot for the average infected effect, defined as (FC + LC)/2 minus Healthy."}
plot_volcano(infected_output, "Infected_vs_Healthy")
```

### Differentially expressed proteins

```{r de-table-infected-vs-healthy}
prepare_de_table(infected_output, "Infected_vs_Healthy") |>
  create_dt(paste0(params$request, "-Infected-over-Healthy-DE-proteins"))
```

## Gene Set Enrichment Analysis

GO Biological Process GSEA follows the pathway-analysis workflow used for this
dataset. Human UniProt accessions are mapped to Entrez IDs with `org.Hs.eg.db`,
then ranked as `sign(log2FC) * -log10(p-value)`. Duplicate Entrez mappings retain
the protein with the largest absolute ranking score. Enrichment uses
Benjamini-Hochberg correction and includes gene sets containing
`r params$gsea_min_gene_set_size` to `r params$gsea_max_gene_set_size` genes.
Positive normalized enrichment scores indicate enrichment toward proteins with
positive MSstats fold changes; negative scores indicate the reverse.

```{r run-go-bp-gsea}
gsea_comparison_labels <- c(
  three_condition_contrasts$label,
  "Infected_vs_Healthy"
)

gsea_results <- purrr::set_names(gsea_comparison_labels) |>
  purrr::map(function(comparison_label) {
    run_go_bp_gsea(
      all_comparison_output,
      comparison_label = comparison_label,
      min_gene_set_size = params$gsea_min_gene_set_size,
      max_gene_set_size = params$gsea_max_gene_set_size,
      pvalue_cutoff = params$gsea_pvalue_cutoff,
      dot_show_categories = params$gsea_dot_show_categories,
      tree_show_categories = params$gsea_tree_show_categories,
      tree_label_wrap = params$gsea_tree_label_wrap
    )
  })

gsea_run_summary <- summarize_gsea_runs(gsea_results)
gsea_run_summary_file <- write_results(gsea_run_summary, "gsea_go_bp_run_summary.csv")

gsea_combined_table <- purrr::imap_dfr(gsea_results, function(result, comparison_label) {
  result$table |>
    mutate(Comparison = comparison_label, .before = 1)
})
gsea_combined_file <- write_results(gsea_combined_table, "gsea_go_bp_all_comparisons.csv")

gsea_result_files <- purrr::imap_chr(gsea_results, function(result, comparison_label) {
  write_results(
    result$table,
    paste0("gsea_go_bp_", tolower(comparison_label), ".csv")
  )
})

knitr::kable(
  gsea_run_summary,
  caption = "GO Biological Process GSEA status and result counts by comparison."
)
```

### GSEA Dot Plots

<span id="fig-gsea-dotplots"></span>

```{r fig-gsea-dotplots, results='asis', fig.width=16, fig.height=10, fig.cap="GO Biological Process GSEA dot plots for each comparison. Point size represents gene count and color represents the adjusted p-value."}
for (comparison_label in names(gsea_results)) {
  result <- gsea_results[[comparison_label]]
  cat("#### ", format_comparison_label(comparison_label), "\n\n", sep = "")
  if (!is.null(result$dotplot)) {
    print(result$dotplot)
  } else {
    cat("*", result$status, "*\n\n", sep = "")
  }
  cat("\n\n")
}
```

### GSEA Tree Plots

<span id="fig-gsea-treeplots"></span>

Tree plots group enriched GO terms by pairwise term similarity, making redundant
or closely related biological processes easier to recognize.

```{r fig-gsea-treeplots, results='asis', fig.width=18, fig.height=14, fig.cap="GO Biological Process similarity trees for each comparison; branch colors group related enriched pathways."}
for (comparison_label in names(gsea_results)) {
  result <- gsea_results[[comparison_label]]
  cat("#### ", format_comparison_label(comparison_label), "\n\n", sep = "")
  if (!is.null(result$treeplot)) {
    print(result$treeplot)
  } else {
    cat("*Tree plot unavailable. ", result$status, "*\n\n", sep = "")
  }
  cat("\n\n")
}
```

### GSEA Results Tables

`NES` is the normalized enrichment score, `p.adjust` is the
Benjamini-Hochberg-adjusted p-value, and `core_enrichment` lists the leading-edge
genes driving the enrichment score.

```{r gsea-results-tables, results='asis'}
invisible(DT::datatable(NULL))
for (comparison_label in names(gsea_results)) {
  result <- gsea_results[[comparison_label]]
  cat("#### ", format_comparison_label(comparison_label), "\n\n", sep = "")
  if (nrow(result$table) > 0) {
    table_widget <- DT::datatable(
      result$table,
      rownames = FALSE,
      caption = htmltools::tags$caption(
        style = "caption-side: top; text-align: left; font-weight: bold;",
        paste(
          "GO Biological Process GSEA results:",
          format_comparison_label(comparison_label)
        )
      ),
      extensions = "Buttons",
      options = list(
        dom = "lBfrtip",
        buttons = list(
          "copy",
          "csv",
          list(
            extend = "excel",
            filename = paste0(
              params$request, "-GSEA-",
              stringr::str_replace_all(tolower(comparison_label), "_", "-")
            ),
            title = paste(
              "GO Biological Process GSEA results:",
              format_comparison_label(comparison_label)
            )
          )
        ),
        pageLength = 10,
        lengthMenu = list(c(10, 25, 50, -1), c("10", "25", "50", "All")),
        scrollX = TRUE
      )
    ) |>
      DT::formatSignif(
        intersect(c("enrichmentScore", "NES", "pvalue", "p.adjust", "qvalue"), names(result$table)),
        digits = 3
      )
    print(htmltools::tagList(table_widget))
  } else {
    cat("*", result$status, "*\n\n", sep = "")
  }
  cat("\n\n")
}
```


## Power and Detectability

Power diagnostics are calculated from the MSstats contrast estimates, standard
errors, and degrees of freedom. The prospective-style calculation asks whether
the current design and observed variability would have `r scales::percent(params$power_target)`
power to detect a target effect of `r abs(params$power_effect_log2fc)` log2 fold
change at a nominal alpha of `r params$power_alpha`. The observed-effect
post-hoc power calculation uses the measured log2 fold change, so it should be
read as a detectability diagnostic alongside the p-values rather than as
independent evidence.

```{r power-settings}
tibble(
  setting = c(
    "Nominal alpha for power calculations",
    "FDR threshold used for detected proteins",
    "Target power",
    "Target absolute log2 fold-change",
    "Target linear fold-change"
  ),
  value = c(
    params$power_alpha,
    params$power_alpha,
    params$power_target,
    abs(params$power_effect_log2fc),
    round(2^abs(params$power_effect_log2fc), 3)
  )
) |>
  knitr::kable(caption = "Power-analysis settings used for all comparisons.")
```

```{r power-diagnostics}
power_diagnostics <- add_power_metrics(
  all_comparison_output,
  alpha = params$power_alpha,
  target_power = params$power_target,
  effect_log2fc = params$power_effect_log2fc
)

power_diagnostics_file <- write_results(power_diagnostics, "msstats_power_diagnostics.csv")

power_summary <- summarize_power_metrics(power_diagnostics)
power_summary |>
  knitr::kable(
    digits = 2,
    caption = "Power and minimum-detectable-effect summary by comparison."
  )
```

```{r power-detection-status}
power_diagnostics |>
  count(.data$Label, .data$power_detection_status, name = "proteins") |>
  group_by(.data$Label) |>
  mutate(percent = 100 * .data$proteins / sum(.data$proteins)) |>
  ungroup() |>
  arrange(.data$Label, desc(.data$proteins)) |>
  knitr::kable(
    digits = 2,
    caption = "Protein counts by power and differential-detection status."
  )
```

```{r fig-power-target-effect, fig.width=10, fig.height=6, fig.cap="Estimated power to detect the configured target log2 fold change across proteins in each comparison."}
power_diagnostics |>
  filter(is.finite(.data$target_log2fc_power)) |>
  ggplot(aes(x = .data$Label, y = .data$target_log2fc_power, fill = .data$Label)) +
  geom_violin(trim = TRUE, alpha = 0.65, show.legend = FALSE) +
  geom_boxplot(width = 0.18, outlier.alpha = 0.15, show.legend = FALSE) +
  geom_hline(yintercept = params$power_target, linetype = "dashed", color = "grey25") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    x = NULL,
    y = paste0("Power to detect |log2FC| = ", abs(params$power_effect_log2fc)),
    title = "Power for the target effect size by comparison"
  ) +
  theme_bw()
```

```{r fig-minimum-detectable-effect, fig.width=10, fig.height=6, fig.cap="Estimated minimum absolute log2 fold change required to attain the target power for each comparison."}
power_diagnostics |>
  filter(is.finite(.data$detectable_log2fc_at_target_power)) |>
  ggplot(aes(x = .data$Label, y = .data$detectable_log2fc_at_target_power,
             fill = .data$Label)) +
  geom_violin(trim = TRUE, alpha = 0.65, show.legend = FALSE) +
  geom_boxplot(width = 0.18, outlier.alpha = 0.15, show.legend = FALSE) +
  geom_hline(yintercept = abs(params$power_effect_log2fc),
             linetype = "dashed", color = "grey25") +
  coord_flip() +
  labs(
    x = NULL,
    y = paste0("Minimum |log2FC| for ", scales::percent(params$power_target), " power"),
    title = "Estimated detectable effect size by comparison"
  ) +
  theme_bw()
```

```{r fig-post-hoc-power, fig.width=10, fig.height=6, fig.cap="Observed-effect power versus absolute log2 fold change, stratified by comparison and FDR detection status."}
power_diagnostics |>
  filter(is.finite(.data$observed_effect_power), is.finite(.data$log2FC)) |>
  ggplot(aes(x = abs(.data$log2FC), y = .data$observed_effect_power,
             color = .data$fdr_detected)) +
  geom_point(alpha = 0.45, size = 1.4) +
  geom_hline(yintercept = params$power_target, linetype = "dashed", color = "grey25") +
  geom_vline(xintercept = abs(params$power_effect_log2fc),
             linetype = "dotted", color = "grey35") +
  facet_wrap(~ Label) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  scale_color_manual(values = c("FALSE" = "grey55", "TRUE" = "#C44E52")) +
  labs(
    x = "Observed absolute log2 fold-change",
    y = "Observed-effect post-hoc power",
    color = paste0("FDR < ", params$power_alpha),
    title = "Observed-effect power and detected proteins"
  ) +
  theme_bw()
```

## Model Issue Summary

MSstats can annotate rows with model-fitting or comparison issues. This summary
counts those flags by comparison so they are visible before exported result
files are used downstream.

```{r model-issue-summary}
add_issue_status <- function(results, analysis_label) {
  results <- as_plain_tibble(results)
  if (!"issue" %in% names(results)) {
    results$issue <- NA_character_
  }

  results |>
    mutate(
      analysis = analysis_label,
      issue = as.character(.data$issue),
      issue = if_else(is.na(.data$issue) | !nzchar(.data$issue),
                      "No issue", .data$issue)
    )
}

comparison_issue_summary <- bind_rows(
  add_issue_status(three_condition_output, "Three-condition"),
  add_issue_status(infected_output, "Infected vs Healthy")
) |>
  count(.data$analysis, .data$Label, .data$issue, name = "proteins") |>
  arrange(.data$analysis, .data$Label, desc(.data$proteins))

knitr::kable(
  comparison_issue_summary,
  caption = "MSstats model and comparison issue counts."
)
```

## Conclusion

```{r conclusion, results='asis'}
render_analysis_conclusion(
  all_comparison_output,
  gsea_combined_table,
  fdr_cutoff = params$volcano_p_cutoff,
  log2fc_cutoff = params$volcano_fc_cutoff,
  power_summary = power_summary,
  target_power = params$power_target,
  power_effect_log2fc = params$power_effect_log2fc
)
```

```{=html}
<script>
document.addEventListener("DOMContentLoaded", function () {
  const conclusion = document.getElementById("conclusion");
  const methods = document.getElementById("methods");
  if (conclusion && methods && methods.parentNode) {
    methods.parentNode.insertBefore(conclusion, methods);
  }

  const tocLink = document.querySelector('#TOC a[href="#conclusion"]');
  if (tocLink) {
    const tocItem = tocLink.closest("li");
    const tocList = tocItem ? tocItem.parentNode : null;
    if (tocItem && tocList) tocList.insertBefore(tocItem, tocList.firstChild);
  }

  document.addEventListener("click", function (event) {
    const link = event.target.closest('a[href^="#fig-"]');
    if (!link) return;
    const targetId = link.getAttribute("href").slice(1);
    const target = document.getElementById(targetId);
    const tabPane = target ? target.closest(".tab-pane") : null;
    if (!target || !tabPane || tabPane.classList.contains("active")) return;

    const selector =
      '[data-bs-target="#' + tabPane.id + '"], [href="#' + tabPane.id + '"]';
    const tabControl = document.querySelector(selector);
    if (tabControl && window.bootstrap) {
      event.preventDefault();
      bootstrap.Tab.getOrCreateInstance(tabControl).show();
      window.setTimeout(function () {
        target.scrollIntoView({behavior: "smooth", block: "start"});
      }, 180);
    }
  });
});
</script>
```

## Session Info

```{r session-info}
sessionInfo()
```
